---
title: "Exposure Therapy or Cognitive Restructuring for Statistics Anxiety?"
author: "Chen Man Hin"
date: "2026-03-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Libraries

To investigate statistics anxiety, all analyses were conducted in R using the bootnet package for network estimation and the igraph package for network visualisation, analysis and simulation.

```{r}
library(igraph)
library(bootnet) # to compute partial correlation matrix
library(tidyverse) # for data wrangling
```

### Data

The 51-item STARS was partitioned into its two theoretically established domains: the affective sub-domain (items 1-23) and cognitive sub-domain (items 24-51). Partial correlation networks were estimated for both sub-networks, as well as the global network, and subsequently converted into undirected, weighted igraph objects for analysis.

```{r}
# load the data
data <- read.csv('stars-data_132.csv', header = T)
# split into sub-networks
affective <- data |> select(item_1:item_23) # affective sub-network
cognitive <- data |> select(item_24:item_51) # cognitive sub-network
all <- data |> select(item_1:item_51)  # full network
# create partial correlation matrix for each sub-network
net_affective <- estimateNetwork(affective, default = "pcor", 
                                threshold = 'sig') # affective partial correlation matrix
net_cognitive <- estimateNetwork(cognitive, default = "pcor", 
                                threshold = 'sig') # cognitive partial correlation matrix
net_g <- estimateNetwork(all, default = "pcor", 
                         threshold = 'sig') # full network
# convert both to igraph objects for analysis
g_affective <- graph_from_adjacency_matrix(abs(net_affective$graph), 
                                           mode="undirected", weighted = TRUE)
g_cognitive <- graph_from_adjacency_matrix(abs(net_cognitive$graph), 
                                           mode="undirected", weighted = TRUE)
g <- graph_from_adjacency_matrix(abs(net_g$graph), 
                                 mode="undirected", weighted = TRUE)
summary(g_affective)
summary(g_cognitive)
```

### Network Visualisation

To visually inspect the structural differences between the sub-domains, the networks were plotted below.

```{r}
# remove 'i_' and 'it_' for presentation and readability
V(g_affective)$label <- gsub("^(i_|it_)", "", V(g_affective)$name)
V(g_cognitive)$label <- gsub("^(i_|it_)", "", V(g_cognitive)$name)

# plot for affective sub-network
plot(g_affective, 
    vertex.color = "lightpink", 
    edge.color = "mediumpurple", 
    vertex.label.color = "black",
    vertex.size = 16,
    vertex.shape = 'circle',
    main = "Affective Subnetwork (Items 1-23)")

# plot for cognitive sub-network
plot(g_cognitive, 
    vertex.color = "lightpink", 
    edge.color = "mediumpurple", 
    vertex.label.color = "black",
    vertex.size = 16,
    vertex.shape = 'circle',
    main = "Cognitive Subnetwork (Items 24-51)")
```

## Hypothesis 1: Internal Cohesion VS Path Efficiency

**Does the cognitive sub-network exhibit greater internal cohesion, whereas the affective sub-system facilitate greater path efficiency?**

_Internal cohesion_ was tested by comparing macro-level metrics including average node strength, edge density, transitivity and degree assortativity; which mathematically assesses the presence of triads in the sub-network. The cognitive components of statistics anxiety are thus expected to exhibit significantly greater values for internal cohesion, presenting structurally as a clustered core of closed triangles, than affective items.

_Path efficiency_ was evaluated using the Average Shortest Path Length (ASPL), where shorter paths indicate greater transmission efficiency. Accordingly, the affective sub-network is expected to exhibit a shorter ASPL than the cognitive sub-network, facilitating faster contagion.

### Macro-Level Analysis: Internal Cohesion

```{r}
cat("\n--- INTERNAL COHESION ---\n")
# Affective Sub-Network
mean(strength(g_affective)) # average node strength
edge_density(g_affective) # density
transitivity(g_affective, type = "global") # global clustering (transitivity)
assortativity_degree(g_affective) # degree assortativity

# Cognitive Sub-Network
mean(strength(g_cognitive)) # average node strength
edge_density(g_cognitive) # density
transitivity(g_cognitive, type = "global") # global clustering (transitivity)
assortativity_degree(g_cognitive) # degree assortativity
```

### Macro-Level Analysis: Path Efficiency

```{r}
cat("\n--- PATH EFFICIENCY ---\n")
# Affective Sub-Network
mean_distance(g_affective, weights = NA) # average shortest path length

# Cognitive Sub-Network
mean_distance(g_cognitive, weights = NA) # average shortest path length
```

## Hypothesis 2: Network Degradation Simulation

**Is structural disintegration, measured by decrease in global edge density, more efficacious from targeted "keystone" symptom deactivation than random symptom alleviation?**

_Keystone symptoms_ were defined as the top five nodes possessing the highest eigenvector centrality within the global network. To ensure fair comparison, five random nodes were deactivated. 

```{r}
cat("\n--- DEGRADATION SIMULATION (REMOVING 5 NODES) ---\n")
baseline_density <- edge_density(g)

# Simulation A: Targeted Removal (Highest Strength)
ec <- eigen_centrality(g)$vector
targeted <- names(sort(ec, 
                       decreasing = TRUE)[1:5]) # identify top 5 nodes with highest ec 
g_targeted <- delete_vertices(g, targeted) # deactivate these nodes
density_targeted <- edge_density(g_targeted) # new density
difference_targeted <- baseline_density - density_targeted

# Simulation B: Random Removal
set.seed(1)
random <- V(g)$name[sample(1:gorder(g), 5, replace = FALSE)] # random 5 nodes
g_random <- delete_vertices(g, random) # deactivate these nodes
density_random <- edge_density(g_random) # new density
difference_random <- baseline_density - density_random
```

### Exploratory Analysis: Micro-Level Symptom Profiling

To translate these macro-level findings into actionable clinical targets, a post-hoc micro-level analysis was performed. Node strength was computed to identify the primary drivers (hubs) of anxiety within each domain, and betweenness centrality was calculated to locate the critical bridges across the network.

```{r}
# Node Strength (Central Hubs)
strength_affective <- strength(g_affective)
strength_cognitive <- strength(g_cognitive)
# Betweenness Centrality (Structural Bridges)
betweenness_affective <- betweenness(g_affective, directed = FALSE, 
                                     weights = 1/abs(E(g_affective)$weight))
betweenness_cognitive <- betweenness(g_cognitive, directed = FALSE, 
                                     weights = 1/abs(E(g_cognitive)$weight))
# Top Three of Affective Sub-network
print(head(sort(strength_affective, decreasing = TRUE), 3))
print(head(sort(betweenness_affective, decreasing = TRUE), 3))
# Top Three of Cognitive Sub-network
print(head(sort(strength_cognitive, decreasing = TRUE), 3))
print(head(sort(betweenness_cognitive, decreasing = TRUE), 3))
```
